Adaptive Resolution Simulation: Theoretical Fundations and Open Questions

### 2014-10-27 writeDown/pub.epj/paper.tex
back up the changes to the MSM, adding intro

@@

@@

\subsection{Velocity auto-correlation of water}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.99\textwidth]{figs/system.alanine/all.eps}
  \caption{(a): A schematic plot of the alanine dipeptide molecule,
    and the two dihedral angles $\phi$ and $\psi$ that are used to
    represent the conformations of the molecule. (b) and (c): The free
    energy landscape plotted using variables $\phi$ and $\psi$.  (b)
    is for full atomistic reference simulation, while (c) is for
    AdResS simulation.}
  \label{fig:ala}
\end{figure}

The alanine dipeptide molecule (see Fig.~\ref{fig:ala}~(a)) is usually
used as a testing case for computational method in simulating real
biomolecular system.
The system presents a very clear time-scale seperation: There are fast
motions like bond and angle vibration that happens on the time-scale of tens of femtoseconds, while
there are slow conformational changes that happens on the time-scale of hundreds of picoseconds.
The dihedral angles $\phi$ and $\psi$ (see
Fig.~\ref{fig:ala}~(a)) are usually chosen as slow variables that  describles
the molecular conformation and its transitions of the moelcules, which are of
general biological interest.

The free energy defined as the logrithm of equilibrium probability density over
$\phi$--$\psi$ space:
\begin{align}
  F(\phi,\psi) = -k_BT \ln p_\equi(\phi,\psi)
\end{align}
is plotted in Fig.~\ref{fig:ala}~(b) and (c). The AdResS free energy is
in good consistency with the full atomistic reference system. This is expected
at least in equilibrium, because as discussed in Sec.~\ref{sec:statistical}, the
AdResS sampling approximates the atomistic grand-canonical ensemble.

In pratice, not only the equilibrium probability density if of interest,
sometimes how the system relax to equilibrium is also of interest.
A direct analysis of the MD trajectories for these properties is not easy.
% Or equivalently
% at which timescale the slow dynamics  happens and from which conformation to which it corresponds to.
One possible solution is the Markov state model (MSM), which
approximate the original high-dimensional dynamics by a finite-state
Markov process (also called reduced dynamics),
and the original time-dependent probability density define on the phase
space $p_t(\vect x)$ is approximated by a  temporal discrete $\vect p_{k\tau}$ probability defined on the finite-state space.
The step of the temporal discretization $\tau$ is called the lag-time.
We always assume the reduced
dynamics is Markovian, although uaually it is not so in practice, and may become one of the main sources of the error.
It would be lengthy to discuss the topic here, and the interested readers are refereed to Ref.~\cite{}.
Due to the Markovianity of the reduced dynamics, the
probability at time $k\tau$ is given by
\begin{align}
\vect p_{k\tau} = \mathcal Q^k(\tau)\circ \vect p_0,
\end{align}
where $\mathcal Q(\tau)$ is the transfer operator
associated to the reduced dynamics. In equilibrium, the transfer operator is time-homogeneous.
Due to the Markovianity, for any $k$, we have
$p_{t+k\tau}(\vect x) = \mathcal Q^k(\tau)\circ p_t (\vect x)$.
If the propagator is irreducible and aperiodic, than according to Perron-Frobenius theorem, the maximum
eigenvalue is single and equal to 1.
If the dynamics
is further assumed to be reversible, then all eigenvalues of the propagator are
real valued.
The first $m$ largest eigenvalues are denoted by $1 = \lambda_1 > \lambda_2 \geq \lambda_3 \geq \cdots \geq \lambda_m$, and
the corresponding eigenvetors are denoted by $\Phi_1, \Phi_2, \cdots, \Phi_m$. The leading eigenvector is nothing but the
equilibrium probability that satisfies $\Phi_1 = \mathcal Q(\tau)\circ \Phi_1$.
The time evolution of the reduced probability density can be expaned by 
\begin{align}\label{eqn:expan}
  \vect p_{k\tau} = \Phi_1 + \sum_{i=2}^m e^{-\frac{k\tau}{t_i}} \,a_i\Phi_i + \textrm{fast decaying dynamics},
\end{align}
where $t_i$ is the timescale that corresponds to the $i$th eigenvalue:
\begin{align}\label{eqn:time-scale}
  t_i  = -\frac{\tau}{\log\lambda_i}, \quad i=2, \cdots, m,
\end{align}
and $a_i$ are constant that are determined by the initial state. As can be seen from Eq.~\eqref{eqn:expan}, the
time-dependent probability  converges to the equilibrium probability as time goes to infinity.
The timescale of the slow part of the relaxation is govenerned by the leanding eigenvectors through Eq.~\eqref{eqn:time-scale},
while the conformational change corresponds to each timescale is indicated by the eigenvector.
We assume a gap in the spectrum between $\lambda_m$ and the rest of the spectrum $\lambda_{m+1}, \cdots$,
therefore, the dynamics of the rest of the spectrum decays out very fast, and is not of particular interest.

The accuracy analysis of the MSM can be find in literature e.g.~\cite{}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/system.alanine/fig-eig-val.png}
  \caption{The non-trivial leading timescale calculated from the
    Markov state modeling for the system of alanine dipeptide
    dissolved in water. In this figure, $t_2$ $t_3$ and $t_4$
    correspond to the first, second and third non-trivial leading
    timescale, respectively. The horizontal axis $\tau$ is the
    lag-time used to build the Markov state model. For a sufficient
    large $\tau$ the values of the timescales should be independent on
    the lag-time. The shadowing regions around the lines indicate the
    statistical uncertainty of the result. }
  \label{fig:ala-msm-t}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-2.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-3.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-4.eps}\\
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-2.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-3.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-4.eps}\\
  \caption{The eigenvectors corresponding to the first three leading timescales.
    The first row presents the eigenvectors of the full atomistic reference, while the second
    row presents those of the AdResS simulation. From left to right $\Phi_2$, $\Phi_3$ and $\Phi_4$ that correspond to
    timescales $t_2$, $t_3$ and $t_4$ are given (see also Fig.~\ref{fig:ala-msm-t}). The magnitude of the eigenvectors are
    denoted by the color. 
    The conformational
    changes indicated by the eigenvectors are noted on the plots.}
  \label{fig:ala-msm-v}
\end{figure}

Here we present the eigenvalues and eigenvectors in
Fig.~\ref{fig:ala-msm-t} and \ref{fig:ala-msm-v}, respectively. The
eigenvalue should be $\tau$ independent if the MSM is of good
accuracy. From the Fig.~\ref{fig:ala-msm-t}, the full atomistic
reference and the AdResS result are consistent within the statistical
uncertainty.  The eigenvectors of the AdResS are also consistent with
those of the full atomistic reference. Although calculated from the
equilibrium MD simulations, the leading timescales and the
corresponding eigenvectors are not the equilibrium properties, because
the transition probability (or the propagator) are actually
time-correlation functions, which are called dynamical properties in
this context. A good reproduction of the equilibrium properties does
not necessary implies a reproduction of the dynamical properties.
Although currently there is still no theoretical answer why AdResS is
so good in reproducing the dynamical properties, the provided
numercial example suggests that AdResS is promising in the applications
beyond equilibrium.

>>>>

\newcommand{\confa}[0]{{\alpha_{\textrm{R}}}}
\newcommand{\confb}[0]{{\textrm{C}7_{\textrm{eq}}}}
\newcommand{\confc}[0]{{\alpha_{\textrm{L}}}}
\newcommand{\confd}[0]{{\textrm{C}7_{\textrm{ax}}}}

@@

\section{Introduction}

Adaptive resolution simulation (AdResS)~\cite{praprotnik2005adaptive,praprotnik2006adaptive,praprotnik2007adaptive,praprotnik2008multiscale,praprotnik2011statistical,fritsch2012adaptive} is a concurrent multiscale simulation method for molecular system.
The terminology ``concurrent multiscale simulation'' means that
part of the system is simulated with higher resolution molecular models for the
purpose of accuracy, and \emph{in the meanwhile} the rest of the system is
simulated with lower resolution models for the purpose of saving computational cost.
The AdResS is, therefore, suitable for the systems that different physcial
phenomena happen concurrently at different time and length scales. For example,
the structure of a protein and its solvation shell
should be resolved at the atomistic level, while far away form the protein,
the only the hydrodynamic properties are of interest, which can be studied with satisfactory
accuracy with very coarse-grained or even continum
models~\cite{zavadlav2014adaptive,zavadlav2014adaptive1,delgado2009coupling}.
Similar  
techniques sharing the ``concurrent'' idea of multiscale simulation can be found in, for example, Ref.~\cite{ensing2007energy,heyes2010thermodynamic,shi2006mixed,shen2014resolution,nielsen2010adaptive}.

Traditionally the

@@

\subsection{Velocity and dipole auto-correlation of water}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.99\textwidth]{figs/system.alanine/all.eps}
  \caption{(a): A schematic plot of the alanine dipeptide molecule,
    and the two dihedral angles $\phi$ and $\psi$ that are used to
    represent the conformations of the molecule. (b) and (c): The free
    energy landscape plotted using variables $\phi$ and $\psi$.  (b)
    is for full atomistic reference simulation, while (c) is for
    AdResS simulation. The metastable conformations coded by $\confa$, $\confb$, $\confc$ and $\confd$
    are noted on plot (b).}
  \label{fig:ala}
\end{figure}

The alanine dipeptide molecule (see Fig.~\ref{fig:ala}~(a)) is usually
used as a testing case for evaluating the new computational method in simulating 
biomolecular system (see e.g.~Ref.~\cite{apostolakis1999calculation,chodera2007automatic,kaminsky2007force,gfeller2007complex}).
The system presents a very clear timescale seperation: There are fast
motions like bond and angle vibration that happens in tens of femtoseconds, while
there are slow conformational transformations that happens in hundreds of picoseconds.
The dihedral angles $\phi$ and $\psi$ (see
Fig.~\ref{fig:ala}~(a)) are usually chosen as slow variables that  describles
the molecular conformation and its transitions of the moelcules, which are of
general biological interest.

The free energy defined as the logrithm of the projected equilibrium probability density on the
$\phi$--$\psi$ space:
\begin{align}
  F(\phi,\psi) = -k_BT \ln p_\equi(\phi,\psi).
\end{align}
The free energy of the full atomistic reference system is showed in
Fig.~\ref{fig:ala}~(b), which clearly presents the metastablity in the
system: several highly populated regions that correspond to metastable
conformations are seperated by lowly populated regions that are
identified as transition regions.
The AdResS free energy is in good
consistency with the full atomistic reference system, and
reproduces the metastable conformations and their populations
with satisfactory accuracy.
This is expected in equilibrium, because as discussed in
Sec.~\ref{sec:statistical}, the AdResS sampling approximates the
atomistic grand-canonical ensemble.

In pratice, not only the equilibrium properties like the free energy are of interest,
but also the dynamical properties like the conformational transitions and the corresponding timescales
are of highly biological importance.
% sometimes how the system relaxes to equilibrium is also of interest,
% e.g.~
% A direct analysis of the MD trajectories for these properties is not easy.
% Or equivalently
% at which timescale the slow dynamics  happens and from which conformation to which it corresponds to.
When the system presents metastablity,
one powerful tool for this purpose
is the Markov state model (MSM)~\cite{prinz2011probing,prinz2011markov,schuette2011markov}, which
approximate the original high-dimensional dynamics by a finite-state reduced
Markov process.
The accuracy analysis regarding  this approximation is well-founded~\cite{sarich2010approximation,djurdjevac2012estimating}.
The MSM has found successfully applications in various fileds, for example studying complex protein systems
(two of the hundreds examples are Ref.~\cite{noe2009constructing,kohlhoff2014cloud}).

In a standard MSM approach, the full phase space is firstly deomposed in to a union of finite number of states
(not necessarily identical to the metastable conformations),
then
the original time-dependent probability density define on the phase
space ($p_t(\vect x)$) is approximated by a temporal discrete $\vect p_{k\tau}$ probability defined on the finite-state space.
The step of the temporal discretization $\tau$ is called the lag-time.
The reduced
dynamics is always assumed to be Markovian,
although uaually it is not,  and in practice it becomes one of the main sources of the error.
% It would be lengthy to discuss the topic here, and the interested readers are refereed to Ref.~\cite{}.
% Due to the Markovianity of the reduced dynamics, the
Under this assumption, the original continuous dynamics is approximated by the Markov processes governed by
% probability at time $k\tau$ is given by
\begin{align}
\vect p^\top_{k\tau} = \vect p^\top_0 \cdot \vect T^k(\tau),
\end{align}
where $\vect T(\tau)$ is the transition maxtix that
associates to the reduced dynamics. The matrix element, for example $T_{ij}$
is defined as the probability of the system being in state
$j$ at $\tau$,  provided the system being in state $i$ at time 0.
% The ``$\cdot$'' denotes the matrix-vector multiplication.
In equilibrium, the transition matrix is time-homogeneous, i.e.~$\vect T(\tau)$ does not depend on time $t$.
% Due to the Markovianity, for any $k$, we have
% $p_{t+k\tau}(\vect x) = \mathcal Q^k(\tau)\circ p_t (\vect x)$.
If the transition matrix is irreducible and aperiodic, than according to Perron-Frobenius theorem, the maximum
eigenvalue is reached and single. Due to the stochaticity of the matrix, the maximal eigenvalue is equal to 1.
If the dynamics
is further assumed to be reversible, then all eigenvalues of the transition are
real valued.
The first $m$ largest eigenvalues are denoted by $1 = \lambda_1 > \lambda_2 \geq \lambda_3 \geq \cdots \geq \lambda_m$, and
the corresponding left eigenvetors are denoted by $\Phi_1, \Phi_2, \cdots, \Phi_m$.
The leading eigenvector is non-negative, and satisfies $\Phi^\top_1 = \Phi^\top_1\cdot\vect T(\tau)$.
With proper normalization the leanding eigenvector is identical to the 
equilibrium probability density.
At inifinitely long time, the probability density $p_{k\tau}$ converges to the equilibrium probability.
The speed of the convergence is characterized by the non-trivial leading timescales that are
determined by the leading eigenvalues:  $t_i  = -{\tau}/{\log\lambda_i}, \ i=2, \cdots, m$.
Since the eigenvectors are orthorgonal and the first eigenvectors is non-negative, the reset
of the eigenvectors have a vanished sum over all elements: Their
value is positive at some states and negative at some other states.
The different signs in the eigenvectors denote the conformational transition corresponding to the
timescale $t_i$.

% If the leading eigenvector is normalized to 1.
% The  reduced probability

% The time evolution of the reduced probability density can be expaned by 
% \begin{align}\label{eqn:expan}
%   \vect p_{k\tau} = \Phi_1 + \sum_{i=2}^m e^{-\frac{k\tau}{t_i}} \,a_i\Phi_i + \textrm{fast decaying dynamics},
% \end{align}
% where $t_i$ is the timescale that corresponds to the $i$th eigenvalue:
% \begin{align}\label{eqn:timescale}
%   t_i  = -\frac{\tau}{\log\lambda_i}, \quad i=2, \cdots, m,
% \end{align}
% and $a_i$ are constant that are determined by the initial state. As can be seen from Eq.~\eqref{eqn:expan}, the
% time-dependent probability  converges to the equilibrium probability as time goes to infinity.
% The timescale of the slow part of the relaxation is govenerned by the leanding eigenvalues through Eq.~\eqref{eqn:timescale},
% while the conformational change corresponds to each timescale is indicated by the eigenvector.
% We assume a gap in the spectrum between $\lambda_m$ and the rest of the spectrum $\lambda_{m+1}, \cdots$,
% therefore, the dynamics of the rest of the spectrum decays out very fast, and is not of particular interest.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{figs/system.alanine/fig-eig-val.png}
  \caption{The non-trivial leading timescale calculated from the
    Markov state modeling for the system of alanine dipeptide
    dissolved in water. In this figure, $t_2$ $t_3$ and $t_4$
    correspond to the first, second and third non-trivial leading
    timescale, respectively. The horizontal axis $\tau$ is the
    lag-time used to build the Markov state model. For a sufficient
    large $\tau$ the values of the timescales should be independent on
    the lag-time. The shadowing regions around the lines indicate the
    statistical uncertainty of the result. }
  \label{fig:ala-msm-t}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-2.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-3.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-atom-4.eps}\\[-.4cm]
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-2.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-3.eps}
  \includegraphics[width=0.3\textwidth]{figs/system.alanine/fig-eig-vec-adress-4.eps}\\
  \caption{The eigenvectors corresponding to the first three leading timescales computed using lag-time $20$~ps.
    The first row presents the eigenvectors of the full atomistic reference, while the second
    row presents those of the AdResS simulation. From left to right $\Phi_2$, $\Phi_3$ and $\Phi_4$ that correspond to
    timescales $t_2$, $t_3$ and $t_4$ are given (see also Fig.~\ref{fig:ala-msm-t}). The magnitude of the eigenvectors are
    denoted by the color. 
    The conformational
    changes indicated by the eigenvectors are noted on the plots.}
  \label{fig:ala-msm-v}
\end{figure}

In the example of alanine dipeptide dissoved in water, the states in the MSM
is defined by uniformly dividing the $\phi$-$\psi$ space in to $20\times 20$ bins.
Each bin presents a state in MSM, and the transition matrix is sampled by equilibrium
MD simulations with the local thermostating scheme.
The MSM and the leading eigen properties are built and analyzed by the software EMMA~\cite{senne2012emma}.
The resulting eigenvalues and eigenvectors of AdResS are compared with the full atomistic reference in
Fig.~\ref{fig:ala-msm-t} and \ref{fig:ala-msm-v}. The
eigenvalues should be $\tau$ independent if the MSM is of good
accuracy, and from Fig.~\ref{fig:ala-msm-t} they are so for $\tau \geq 20$~ps.
Also from the Fig.~\ref{fig:ala-msm-t}, the full atomistic
reference and the AdResS result are consistent within the statistical
uncertainty.
From Fig.~\ref{fig:ala-msm-v}, the conformational transitions corrsponds
to $t_2$, $t_3$ and $t_4$ are $\{\confa \} \leftrightarrow \{\confc,\confd\}$,
$\{\confa, \confc \} \leftrightarrow \{\confb\}$ and $\{\confc \} \leftrightarrow \{\confd\}$, respectively.
The eigenvectors and the indicated conformational transitions of the AdResS are also consistent with
those of the full atomistic reference. Although calculated from the
equilibrium MD simulations, the leading timescales and the
corresponding eigenvectors are not the equilibrium properties, because
the transition probability (elements of the transition matrix) are actually
time-correlation functions of the characteristic functions, which are called dynamical properties in
this context. A good reproduction of the equilibrium properties does
not necessary implies a reproduction of the dynamical properties.
Although currently there is still no theoretical answer why AdResS works,
the provided numercial example suggests that AdResS being promising in the applications beyond equilibrium,
and  worth further theoretical studies.

\section{Open problems of the adaptive resolution simulation}
